Survival in chessland

### 2018-12-15 chess/papers/survival.tex
Generate and import survival maps for each piece

\documentclass[twocolumn]{article}
\usepackage[top=1.1in, left=0.85in, right=0.85in]{geometry}

@@

@@

To compute the chances for survival, I legally acquired

@@

>>>>

\documentclass[twocolumn]{article}
\usepackage[top=0.5in, left=0.45in, right=0.45in, bottom=0.5in]{geometry}

@@

% \usepackage{chessfs}
\usepackage{adjustbox}

% Define black versions of pieces for inline use. Gross, but it works.
\newcommand{\Pawn}[1][1.3ex]{%
\adjustbox{Trim=4.3pt 2.6pt 4.3pt 0pt,width=#1,margin=0.2ex 0ex 0.2ex 0ex}{\BlackPawnOnWhite}%
}%
\newcommand{\Rook}[1][1.58ex]{%
\adjustbox{Trim=3.2pt 2.2pt 3.2pt 0pt,width=#1,raise=0ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackRookOnWhite}%
}%
\newcommand{\Knight}[1][1.85ex]{%
\adjustbox{Trim=2.3pt 2.35pt 2.5pt 0pt,width=#1,raise=-0.03ex,margin=0.14ex 0ex 0.14ex 0ex}{\BlackKnightOnWhite}%
}%
\newcommand{\Bishop}[1][1.79ex]{%
\adjustbox{Trim=2.3pt 2pt 2.3pt 0pt,width=#1,raise=-0.12ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackBishopOnWhite}%
}%
\newcommand{\Queen}[1][2.05ex]{%
\adjustbox{Trim=1.2pt 2.2pt 1.2pt 0pt,width=#1,raise=-0.08ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackQueenOnWhite}%
}%
\newcommand{\King}[1][1.95ex]{%
\adjustbox{Trim=2pt 2pt 2pt 0pt,width=#1,raise=-0.06ex,margin=0.13ex 0ex 0.13ex 0ex}{\BlackKingOnWhite}%
}%

@@

To compute the chances for survival, I legally acquired 506,000,416
chess games from lichess.org. This is all of the standard variant,
rated games from Jan 2013 to November 2018, in any time format. Only
games that are completed and valid are included (about 200,000 games
did not meet this criteria). The total data size is 875 gigabytes, so
processing these took some care for efficiency and parallelism.

Other than that, I simply parsed and simulated each game. For each
of the 32 pieces in the starting position, I tracked its current
location, and whether it is alive; multiple dead pieces could
occupy the same square. At the end of the game, one of the kings is
killed if his side has lost.

For a piece, there is a single factual survival rate in these games,
given just by $\frac{\mathrm{num~survived}}{\mathrm{num~games}}$. What
we're really interested in, however, is estimating the underlying true
survival probability for each piece. In order to do this with
reasonable efficency, we divide the games into 32 separate buckets,
and count statistics separately for each. From these samples we can
then estimate variance, for example. The games are bucketized by a
deterministic hash of the White player's username. This way, if there
exist some players who are highly unusual (perhaps automated
accounts), their games are grouped together and pessimistically
represented in the variance estimate. This also helps account for
different opening preferences; the chosen opening certainly affects
the survival chances.

\begin{figure}[htb]
  \begin{center}
    % To generate this pdf, I ran maketable to make
    % piece-survival.svg, then printed to PDF using Chrome, then
    % cropped it in Illustrator. Illustrator can open the SVG but
    % doesn't have the unicode chess pieces in any font I could find.
    \includegraphics[width=0.95 \linewidth]{piece-survival-export}
  \end{center}\vspace{-0.1in}
  \caption{Survival probabilities for each of the 32 pieces in standard
    chess, in 500 million games. The number is the mean survival rate
    across all samples. The vertical position of the dot is this mean
    rate, with a line drawing the span between the smallest and largest
    sample bucket (this is usually a very tiny range).
    Horizontal position is purely presentational, to avoid overlap.}
  \label{fig:piece-survival}
\end{figure}

The basic survival chances appear in Figure~\ref{fig:piece-survival}.
Indeed, many pieces are more likely to survive than the kings. Even as
black, the extremal pawns (\pawn a and \pawn h) have over a 70\%
survival rate. Across the board, the survival chances for a white
piece and its black twin are similar, usually with a small edge to
white. Notable exceptions are the \knight g (the overall most doomed
piece), and both white bishops, which die more than their black
counterparts. The \pawn e is vastly more dead than \Pawn e. Note
somewhat satisfyingly that the \pawn c has the highest variance; this
was the most controversial among the drinking buddies
(Section~\ref{sec:XXXbuddies}). Note that \pawn c is the sacrificed pawn
in the popular Queen's Gambit (1. d4 d5 2. c4); this may be a good
example of a piece that has very different survival rates in different
opening preferences. Since the variance is otherwise very low, I did
not attempt to

Despite my impression that many games end in a draw, ties are actually
rare in the lichess database. In January 2018, only 3.8\% of games
were drawn;
% (8922275 white, 8329221 black, 682042 draw)
as a result, the survival chances for the kings are both close to
50\%.

\section{Safest spaces}

The fate of each piece is to either survive or die, and it does so on
one of the 64 squares. With the same replay of the 500 million games
I also kept statistics on the fates of each piece. If being a chess
piece to the death, and possessing some influence over the

@@

Other than the bishops---which have no legal way to reach half of the
squares---every piece reaches every square in at least a thousand
games. So we have enough samples to have reasonable confidence in our
statistics, even for the most unlikely odysseys. The least mobile
pieces are the pawns, who can technically reach any square by
promoting, but are usually confined to cones emanating from their
start squares. The overall rarest fate is for \pawn f2 to die on the
a7 square, which only happened 1,244 times (however, it survived on
this square 31,438 times). This square is actually reachable without
promoting, but it would need to capture 5 times in order to get
there, which seems quite unlikely! There may even be a hidden
achievement for reaching this square this way!
% TODO: Perhaps could find a game where this happened?!
Aside from pawns, the weirdest fate is for \knight g to die on h1,
which happened 47,307 times. Corners are of course fairly garbage for
the knight, although it is twice as likely to survive on this square.

% There may even be hidden achievements for reaching these squares!

That said, there are characteristic patterns for each of the pieces,
which make sense given their starting positions and patterns of
movement. You could probably guess the piece just by looking at one of
the heat maps below, although---spoiler alert---the piece is just
listed right there and they are in order. In each map, the background
color gives the

\newcommand{\chessmap}[2]{\begin{center} \includegraphics[width=0.65 \linewidth]{#1} {#2} \end{center}\vspace{-0.1in}}

\chessmap{piece0}{\Rook a8}
\chessmap{piece1}{\Knight b8}
\chessmap{piece2}{\Bishop c8}
\chessmap{piece3}{\Queen}
\chessmap{piece4}{\King}
\chessmap{piece5}{\Bishop f8}
\chessmap{piece6}{\Knight g8}
\chessmap{piece7}{\Rook h8}

\chessmap{piece8}{\Pawn a7}
\chessmap{piece9}{\Pawn b7}
\chessmap{piece10}{\Pawn c7}
\chessmap{piece11}{\Pawn d7}
\chessmap{piece12}{\Pawn e7}
\chessmap{piece13}{\Pawn f7}
\chessmap{piece14}{\Pawn g7}
\chessmap{piece15}{\Pawn h7}

\chessmap{piece16}{\pawn a2}
\chessmap{piece17}{\pawn b2}
\chessmap{piece18}{\pawn c2}
\chessmap{piece19}{\pawn d2}
\chessmap{piece20}{\pawn e2}
\chessmap{piece21}{\pawn f2}
\chessmap{piece22}{\pawn g2}
\chessmap{piece23}{\pawn h2}

\chessmap{piece24}{\rook a1}
\chessmap{piece25}{\knight b1}
\chessmap{piece26}{\bishop c1}
\chessmap{piece27}{\queen}
\chessmap{piece28}{\king}
\chessmap{piece29}{\bishop f1}
\chessmap{piece30}{\knight g1}
\chessmap{piece31}{\rook h1}
